---
title: "Untitled"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

```{r}

# I am lading the needed packages at the beginning of the script

library(readr)
library(dplyr)
library(ggplot2)

# An R package to access the World Bank World Development Indicator timeseries

library(WDI)

# An R package for working with country codes

library(countrycode)

```

# Data import

## What a waste!

https://datacatalog.worldbank.org/search/dataset/0039597

The data we use comes from ... \[TODO\].

```{r}

# This code is copied from application exercise week 10

global_waste_data <- read_csv("data/raw_data/country_level_data_0.csv")
```

## World Development Indicators (WDI)

GitHub repo: https://github.com/vincentarelbundock/WDI R Package Documentation: http://vincentarelbundock.github.io/WDI/

```{r}


WDIsearch("GDP per capita ")

gdp_data <- WDI(indicator = "NY.GDP.PCAP.CD", start = 2019, end = 2019) 


```

# Data tidying

## Waste data

```{r}

# This code is copied from application exercise week 10

global_waste_data_small <- global_waste_data %>% 
  select(country_name, 
         iso3c, 
         income_id, 
         total_msw_total_msw_generated_tons_year, 
         population_population_number_of_people) %>%  
  rename(msw_tons_year = total_msw_total_msw_generated_tons_year,
         population = population_population_number_of_people) 

```

## GDP data

```{r}

gdp_data_tidy <- gdp_data %>% 
  as_tibble() %>% 
  rename(gdp_capita = NY.GDP.PCAP.CD)

```

# Data transformation

## Waste data

```{r}

# This code is copied from application exercise week 10

global_waste_data_kg_year <- global_waste_data_small %>% 
  mutate(capita_kg_year = msw_tons_year / population * 1000) %>% 
  mutate(income_id = factor(income_id, 
                            levels = c("HIC", "UMC", "LMC", "LIC"))) %>% 
  # select relevant columns
  select(country_name:income_id, capita_kg_year)

```

## GDP data

```{r}

gdp_data_join <- gdp_data_tidy %>%
  mutate(iso3c = countrycode(iso2c, origin = "iso2c", destination = "iso3c")) %>% 
  filter(!is.na(iso3c)) 

```

## Joining two datasets

```{r}

waste_gdp_data <- global_waste_data_kg_year %>% 
  left_join(gdp_data_join) %>% 
  select(country_name:capita_kg_year, gdp_capita, year)

```

# Data visualisation

## Aesthetic mappings

```{r}

ggplot(data = waste_gdp_data,
       mapping = aes(x = gdp_capita, 
                     y = capita_kg_year)) +
  geom_point()

```

```{r}

ggplot(data = waste_gdp_data,
       mapping = aes(x = gdp_capita, 
                     y = capita_kg_year,
                     color = income_id,
                     shape = income_id)) +
  geom_point()

```

## Scaling

### Axis scales 

```{r}

library(scales)

ggplot(data = waste_gdp_data,
       mapping = aes(x = gdp_capita, 
                     y = capita_kg_year,
                     color = income_id,
                     shape = income_id)) +
  geom_point() +
  
  # https://scales.r-lib.org/reference/label_log.html
  scale_x_log10(labels = scales::label_log(), limits = c(100, 150000))  +
  scale_y_continuous(breaks = seq(0, 1400, 200), limits = c(0, 1400)) 


seq(from = 0, to = 1400, by = 200)
c(0, 1400)

```

### Color scales

- scale_color_brewer()

```{r}

ggplot(data = waste_gdp_data,
       mapping = aes(x = gdp_capita, 
                     y = capita_kg_year,
                     color = income_id,
                     shape = income_id)) +
  geom_point() +
  
  # https://scales.r-lib.org/reference/label_log.html
  scale_x_log10(labels = scales::label_log(), limits = c(100, 150000))  +
  scale_y_continuous(breaks = seq(0, 1400, 200), limits = c(0, 1400)) +
  scale_color_brewer(type = "qual", palette = "Set2")

```

- scale_color_manual()

```{r}

my_colors <- c("#ED6D37", "#901C6C", "#18A5C5", "#A6CE8F")

ggplot(data = waste_gdp_data,
       mapping = aes(x = gdp_capita, 
                     y = capita_kg_year,
                     color = income_id,
                     shape = income_id)) +
  geom_point(size = 3, alpha = 3/4) +
  geom_smooth(method = "lm", se = FALSE) +
  
  # https://scales.r-lib.org/reference/label_log.html
  scale_x_log10(labels = scales::label_log(), limits = c(100, 150000))  +
  scale_y_continuous(breaks = seq(0, 1400, 200), limits = c(0, 1400)) +
  scale_color_manual(values = my_colors)

```

## Geoms

- defined globally in ggplot

```{r}

ggplot(data = waste_gdp_data,
       mapping = aes(x = gdp_capita, 
                     y = capita_kg_year,
                     color = income_id,
                     shape = income_id)) +
  geom_point(size = 3, alpha = 3/4) +
  geom_smooth(method = "lm", se = FALSE) +
  
  # https://scales.r-lib.org/reference/label_log.html
  scale_x_log10(labels = scales::label_log(), limits = c(100, 150000))  +
  scale_y_continuous(breaks = seq(0, 1400, 200), limits = c(0, 1400)) +
  scale_color_manual(values = my_colors)

```

- defined for a geom only 

```{r}

ggplot(data = waste_gdp_data,
       mapping = aes(x = gdp_capita, 
                     y = capita_kg_year)) +
  geom_point(size = 3, alpha = 3/4, 
             aes(color = income_id,
                 shape = income_id)) +
  geom_smooth(method = "lm", se = FALSE) +
  
  # https://scales.r-lib.org/reference/label_log.html
  scale_x_log10(labels = scales::label_log(), limits = c(100, 150000))  +
  scale_y_continuous(breaks = seq(0, 1400, 200), limits = c(0, 1400)) +
  scale_color_manual(values = my_colors)
```


## 

# Data modeling

# Data communication
